% !TEX root = WaitFreeRingBuffer.tex

The ring buffer is a staple data structure in computer science applications.
In contrast to linked-list and vector data structures, ring buffers use a constant amount of memory.
This property makes it ideal for certain applications such as multimedia, network routing, trading systems, and gaming. 
In some of these cases, concurrent access is ideal or mandatory.
In this paper, we present a new non-blocking ring buffer that provides a wait-free progress guarantee suitable for such applications.

Several non-blocking queue implementations exist in the literature, however, we are not aware of any that provides the wait-free guarantee.
The design of efficient wait-free algorithms is challenging, because each thread operating on the data structure must complete its operation in a finite number of steps. 
This strict guarantee makes it ideal for real-time and mission critical systems, however, the added complexity to achieve this often results in reduced performance. \NOTES{revolutions in data memory (cite memsql)} %THIS NEEDS TO BE MADE BIG IMPACT
% CAN HELP THINGS LIKE MEMSQL

Circumventing such pitfalls, the presented concurrent ring buffer uses a methodology for diffusing contention, which results in a significant increase in performance.
Operating with 64 threads on a 64-core machine, performance results show that our algorithm performs on average X more operations than a coarse-grained locking approach, X more than TBB's concurrent bounded queue, and X more than the cycle queue by Tsigas.